---
title: "Project_1_Olympic"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (dplyr)
library (tidyr)
```
required libraries:
library (dplyr)
library (tidyr)

Unzipping of files was done manually.

## Task 1


```{r all_files}

all_files <- list.files (path = "../data/", pattern = "*.csv", full.names = TRUE)
all_files
df = lapply(all_files, function(i){
  x  = read.csv(i)
  x
})
df = do.call("rbind.data.frame", df)
```
The result is a dataset of 271115 observations of 15 variables.


## Task 2
```{r summary}
summary (df)
```
```{r is.na}
sum(is.na(df))
```

```{r age}
df %>% filter (Age > 80)
unique (df$Age)
```
There are 9 entries in which athlete's age is over 80. For now, I have decided to drop the most obvious mistake - 240 year.
The others participated in art competitions, and I checked the participant of 84 years - he really was that age. 
```{r age.na}
df$Age[df$Age==240] <- NA
```
The same was done with height - 340 was excluded.
```{r height}
df$Height[df$Height==340] <- NA
```
With sex, "G" and "" were changed to be NAs. 
```{r sex}
df$Sex[df$Sex=="G"] <- NA
```
Weight of 7 was excluded as well. The next minimum was around 25, but it was a weight of a 14-year old female gymnast, so it would make sense. 
```{r weight}
df$Weight[df$Weight=="7"] <- NA
```
There were some typos in Games' names, so they were just corrected.
```{r 2004_summer}
df$Games[df$Games=="2004 Summe"] <- "2004 Summer"
df$Games[df$Games=="2000 Su"] <- "2000 Summer"
df$Games[df$Games==""] <- NA
df$Medal[df$Medal==""] <- NA
```
In general, the method of looking for invalid values was using the "unique" command or looking at summaries. Then obvious typos or non-sense values were excluded. Still, it would be possible to miss some of the mistakes in columns such as "Team" which include a lot of unique values.

## Task 3
Age of the youngest male and female athletes at 1992 Olympics:
```{r age_1992}
df %>% filter (Games == "1992 Summer") %>% group_by(Sex) %>% slice(which.min(Age))
```
So it is 12 for a girl and 11 for a boy. 


## Task 4
Mean and sd for the Height variable for male and female athletes:
```{r height_both}
df %>% group_by(Sex) %>% summarise(mean = mean(Height, na.rm = T), sd = sd(Height, na.rm = T))
```
F: mean 168, sd = 8.80
M: mean 179, sd = 9.36

## Task 5
Height for tennis female athletes at 2000 Olympics:
```{r height_tennis}
df %>% filter (Sex == "F") %>% filter (Games == "2000 Summer") %>% filter (Sport == "Tennis") %>% summarise(mean = round(mean(Height, na.rm = T), 1), sd = round(sd(Height, na.rm = T), 1))
```


## Task 6
The heaviest athlete at 2006 Olympics:
```{r heaviest}
df %>% filter (Games == "2006 Winter") %>% slice(which.max(Weight)) %>% select (Sport)
```
He participated in Skeleton.


## Task 7
Women gold medals from 1980 to 2010
```{r gold}
df %>% filter (Year >= 1980 & Year <= 2010) %>% filter (Sex == "F") %>% summarise(
  count = n(),
  gold_count = sum(Medal == "Gold", na.rm = T)
)
```
2249 gold medals. 


## Task 8
John Aalberg participated in Games:
```{r john}
df %>% filter (Name == "John Aalberg") %>% group_by(Year) %>% 
  summarise(n = n())
```
4 events in 1992 and also 4 in 1994.


## Task 9
One athlete could participate in several events, so one athlete was only counted one time:
```{r 15_25}
df %>% filter (Games == "2008 Summer") %>% filter (Age >= 15 & Age < 25) %>% count(Name) %>% summarise(n = n())
```
For age group 15-25 there were 4783 athletes
```{r 25-35}
df %>% filter (Games == "2008 Summer") %>% filter (Age >= 25 & Age < 35) %>% count(Name) %>% summarise(n = n())
```
For age group 25-35 there were 5376 athletes
```{r 35-45}
df %>% filter (Games == "2008 Summer") %>% filter (Age >= 35 & Age < 45) %>% count(Name) %>% summarise(n = n())
```
For age group 35-45: 630 athletes
```{r 45-55}
df %>% filter (Games == "2008 Summer") %>% filter (Age >= 45 & Age < 55) %>% count(Name) %>% summarise(n = n())
```
For age group 45-55: 76 athletes.

So the highest number of athletes (5376) is within 25-35 age range, the lowest (76) is for 45-55 age range.


## Task 10

Various sports at 2002 and 1994 Olympics:
```{r olympics_2002}
df %>% filter (Games == "2002 Winter") %>% count(Sport) %>% summarise(n = n())
```
```{r olympics_1994}
df %>% filter (Games == "1994 Winter") %>% count(Sport) %>% summarise(n = n())
```
15 sports at 2002 Olympics and 12 at 1994 Olympics.


## Task 11
```{r winter_top}
df %>% filter (Season == "Winter") %>% filter(!is.na(Medal))%>% group_by(Medal, NOC) %>% 
  summarise(n = n()) %>% arrange(desc(n)) %>% group_by(Medal) %>% slice (1)
```
During winter Olympics, Finland has the most bronze medals, USA - most silver, Canada - most gold.
```{r summer_top}
df %>% filter (Season == "Summer") %>% filter(!is.na(Medal))%>% group_by(Medal, NOC) %>% 
  summarise(n = n()) %>% arrange(desc(n)) %>% group_by(Medal) %>% slice (1)

```
During the summer Olympics, USA has the highest number of medals in each category

## Task 12
Z scores of height:
```{r height_z_scores}
df$Height_z_scores <- scale (df$Height)
```


## Task 13
Additional task 


## Task 14
Compare height, weight and age of men of women. T test was used. There was no need to check the normality as the number of observations was huge. Also, Shapiro Wilk test has a limitation of 5000 observations as its upper border.
```{r height_t}
df_women_winter <- df %>% filter (Season == "Winter") %>% filter (Sex == "F")
df_men_winter <- df %>% filter (Season == "Winter") %>% filter (Sex == "M")
t.test(df_women_winter$Height, df_men_winter$Height)
```
t = -184.8, df = 29342, p-value < 2.2e-16

```{r box_height}
df_winter <- df %>% filter (Season == "Winter")
boxplot(df_winter$Height ~ df_winter$Sex, main="Height in Men and Women", xlab="Sex: Female and Male", ylab="Height" )
```
The differences can be observed on boxplot. 
```{r weight_t}
t.test(df_women_winter$Weight, df_men_winter$Weight)
```
t = -187.85, df = 36296, p-value < 2.2e-16
```{r box_weight}
boxplot(df_winter$Weight ~ df_winter$Sex, main="Weight in Men and Women", xlab="Sex: Female and Male", ylab="Weight" )
```

```{r age_t}
t.test(df_women_winter$Age, df_men_winter$Age)
```
t = -32.239, df = 29455, p-value < 2.2e-16
```{r box_age}
boxplot(df_winter$Age ~ df_winter$Sex, main="Age in Men and Women", xlab="Sex: Female and Male", ylab="Age" )
```

## Task 15
Team and Medal - can they be related?

```{r chisq}
df_no_na <- df %>% filter(!is.na(Medal)) %>% filter(!is.na(Team))
chisq.test(table(df_no_na$Team, df_no_na$Medal))
```

X-squared = 29627, df = 1500, p-value <
2.2e-16



## Conclusion

In this project data on Olympic athletes was analyzed. After initial data extraction and pre-processing, exploratory data analysis was done. Statistical methods - such as t-test, chi-squared - were used to compare differences between samples. 

